Simulations of Ballot Polling Risk-Limiting Audits

### 2023-02-28 providence/sections/intro.tex
Changes to intro and bib

Ballot polling audits proceed as follows. 
\begin{enumerate}
\item The ballot manifest is published. 
\item A first round size \cite{usenix_minerva} is chosen. %\fpo{the sentence is hard to understand}. pv: fixed
\item Ballots on the ballot manifest are sampled uniformly at random, with replacement, using a pseudorandom number generator, which is typically seeded by a natural source of randomness like rolling dice.
\item The physical ballots are found and manually interpreted, recording the manual interpretations. 
\item The stopping condition $\mathcal{A}$, a function of the manual interpretations of the current cumulative sample of ballots $X$, is computed. It outputs: 
\begin{enumerate}
\item \emph{Correct: stop the audit} or 
\item \emph{Undetermined: sample more ballots}. 
\end{enumerate}
Election officials may choose to abort this procedure and go to a full hand count at any time and should have a plan for how to decide whether to do so; we discuss this in more detail below and in section \ref{sec:model}.     
\item If more ballots are to be drawn, the next round size is chosen, and we go back to step 2. 
\end{enumerate}

Round sizes, including the first one, may be computed based on a desired probability of audit completion at the end of the round, and may take into consideration loose estimates of the resources required. For RLAs required by statute or legislation, the successful completion of the RLA (or a full hand count confirming the outcome) is usually necessary before certification\footnote{If an RLA were performed after certification and determined that the outcome was incorrect, there may not be a legal means of changing the outcome and this could significantly impact citizen confidence. For example, till recently, the state of Virginia required RLAs but they were to be performed after certification and could not be used to change an outcome. This was corrected through new legislation passed in April 2022.} and certification deadlines would play a large role in round size and hand count decisions.

@@

\npara{\bf Full Hand Count:} It is worth noting here that if election officials do not have a plan for when they will move to a full hand count, and the election outcome is incorrect, at least a fraction $(1-\alpha)$ of the audits will {\em never} stop, and new rounds will continue to be drawn. See Figure \ref{fig:prov-risk} in section \ref{sec:sims} for the results of 10,000 simulations of \Providence on tied elections (risk limit of 0.1, across election margins of 0.05 and larger in the statewide contests for US President in 2020), each simulation stopping after five rounds. Observe that, for each margin studied, more than 90\% of the audits do not stop after five rounds. We present two example simulations from among the ones reported for the state of Texas to illustrate this idea.

@@

A first round size of 2,217 corresponds to a stopping probability of 0.9 and a risk limit of $\alpha=0.1$ for \Providence (see \cite{usenix_minerva} for a description of how first round size may be computed, \Minerva and \Providence are identical in the first round). The audit stops when the inverse of the \Providence ratio (see section \ref{sec:prov}) is smaller than $\alpha$. We describe two audit simulations using these parameters.

The first simulation assumed that the tally is correct and resulted in 1,138 votes for Trump and 1,054 for Biden (the rest were irrelevant votes). The inverse of the \Providence ratio was $0.047 < \alpha$ and the audit stopped.

The second simulation assumed a tied election and did not stop at the end of five rounds. 
%The design of the \Providence audit (see section \ref{sec:prov}) ensures that the risk of the first round is no more than 0.9 times the risk limit, and the risk expended in the first round is roughly 0.09. 
Table \ref{tab:toy} lists the cumulative round sizes, cumulative votes for both candidates and the inverse \Providence ratio after each draw. Each round size was chosen for a conditional stopping probability of 0.9, given that the audit did not stop so far. For example, this corresponds to an approximate probability of 0.09 that the audit stops in the second round (that it does not stop in the first round and stops in the second), and so on. 
\begin{table}[h]
\centering
%\scriptsize
\begin{tabular}{|crrrc|}
\hline
\hline
Round& Cumulative& Trump & Biden & Inverse  \\
 No. & Round Size &  & & Ratio \\
\hline
\hline
1 & 2,217 & 1,111 & 1,079 & 0.256\\
2 & 5,970 & 2,940 & 2,953 & 4.251\\
3 & 16,685 & 8,281 & 8,171 & 3,150 \\
4 & 35,096 & 17,320 & 17,264 & 380,220,376 \\
5 & 76,979 & 37,943 & 37,868 & $2.5e^{+22}$ \\
\hline
\end{tabular}
\caption{A simulation of a \Providence audit of a tied contest with an announced margin of 0.0566, and $\alpha=0.1$. Each round size is computed for a conditional stopping probability of $0.9$. The stopping condition is Inverse Ratio $\le \alpha$.}
\label{tab:toy}
\end{table}

@@

\npara{\bf The Adversary in an RLA:} The goal of the adversary is to increase the true risk beyond the declared risk limit---that is, make the audit declare an incorrect outcome as correct with a chance larger than the risk limit. An audit is an RLA only if there is a proof of its risk-limiting property. Hence, at the very least, the adversary would need to invalidate an assumption of the proof to obtain:  
$$
Pr[\mathcal{A}(X) 
= \text{Correct} \,|\, H_0] > \alpha
$$

@@

The stopping condition for \Minerva does not take into account the sample obtained in previous rounds, and, in \cite{usenix_minerva}, its risk limit is estimated through weighted averages across multiple rounds, assuming that round sizes do not depend on the previous sample. 
%Further, it assumes that the null hypothesis is a tie. 
Attempting to simplify the proof of \Minerva's risk-limiting property, we were able to define a different audit \Providence. The RLA proof for \Providence does not make an assumption about round sizes.

@@

\subsubsection{Organization} 
Sections \ref{sec:model} provides preliminaries on the \BRAVO and \Minerva audits. Section \ref{sec:prov} describes the \Providence audit, section \ref{sec:sims} the simulations comparing the number of ballots drawn using various ballot polling audits and section \ref{sec:pilot} the use of \Providence in an audit carried out by the Board of Elections of Rhode Island. Section \ref{sec:workload} presents our workload model and describes its use for a ballot polling audit using details of the 2016 US Presidential election in the state of Virginia. Our conclusions, the availability of an audit implementation and acknowledgements may be found in sections \ref{sec:conc}, \ref{sec:avail} and \ref{sec:ack} respectively.

>>>>

Ballot polling audits proceed as follows. 
\begin{enumerate}
\item The ballot manifest is published. 
\item A first round size \cite{usenix_minerva} is chosen. %\fpo{the sentence is hard to understand}. pv: fixed
\item Ballots on the ballot manifest are sampled uniformly at random, with replacement, using a pseudorandom number generator---typically seeded by a natural source of randomness like rolling dice.
\item The physical ballots are found and manually interpreted; the interpretations are recorded. 
\item The stopping condition $\mathcal{A}$, a function of the manual interpretations of the current cumulative sample of ballots $X$, is computed. It outputs: 
\begin{enumerate}
\item \emph{Correct: stop the audit} or 
\item \emph{Undetermined: sample more ballots}. 
\end{enumerate}
Election officials may choose to abort this procedure and go to a full hand count at any time and should have a plan for how to decide whether to do so; we discuss this in more detail below and in section \ref{sec:model}.     
\item If more ballots are to be drawn, the next round size is chosen, and the audit goes back to step 2. 
\end{enumerate}

Round sizes, including the first one, may be computed based on a desired probability of audit completion at the end of the round, and may take into consideration loose estimates of the resources required. For RLAs required by statute or legislation, the successful completion of the RLA (or a full hand count confirming the certified outcome) is usually necessary before certification\footnote{If an RLA were performed after certification and determined that the outcome was incorrect, there may not be a legal means of changing the outcome and this could significantly impact citizen confidence. For example, till recently, the state of Virginia required RLAs but they were to be performed after certification and could not be used to change an outcome. This was corrected through new legislation passed in April 2022.} and certification deadlines would play a large role in round size and hand count decisions.

@@

\npara{\bf Full Hand Count:} It is worth noting here that if election officials do not have a plan for when they will move to a full hand count, and the election outcome is incorrect, at least a fraction $(1-\alpha)$ of the audits will {\em never} stop, and new rounds will continue to be drawn. See Figure \ref{fig:prov-risk} in section \ref{sec:sims} for the results of 10,000 simulations of \Providence on tied elections (risk limit of 0.1, across election margins of 0.05 and larger in the statewide contests for US President in 2020). Observe that, for each margin studied, more than 90\% of the audits do not stop after five rounds. To illustrate this idea, we present two example simulations from among those used to compute the statistics reported in Figure \ref{fig:prov-risk}.

@@

A first round size of 2,217 corresponds to a stopping probability of 0.9 and a risk limit of $\alpha=0.1$ for \Providence (see \cite{usenix_minerva} for a description of how first round size may be computed, \Minerva and \Providence are identical in the first round). If the \Providence ratio is denoted $\omega$, the audit stops when $\omega^{-1} \leq \alpha$, or equivalently (see Definition \ref{def:minervatwo}) if $\omega \geq \alpha^{-1}$. We describe two audit simulations using these parameters.

The first simulation assumed that the tally is correct and resulted in 1,138 votes for Trump and 1,054 for Biden (the rest were irrelevant votes) and the audit stopped because: \[ \omega ^{-1} = 0.047 < \alpha \]

The second simulation assumed a tied election and did not stop at the end of five rounds. 
%The design of the \Providence audit (see section \ref{sec:prov}) ensures that the risk of the first round is no more than 0.9 times the risk limit, and the risk expended in the first round is roughly 0.09. 
Table \ref{tab:toy} lists the cumulative round sizes, cumulative votes for both candidates and the inverse \Providence ratio after each draw. Each round size was chosen for a conditional stopping probability of 0.9, given that the audit did not stop so far. For example, this corresponds to an approximate probability of 0.09 that the audit stops in the second round (that it does not stop in the first round and stops in the second), and so on. 
\begin{table}[h]
\centering
%\scriptsize
\begin{tabular}{|crrrc|}
\hline
\hline
Round& Cumulative& Trump & Biden & $\omega^{-1}$  \\
 No. & Round Size &  & &  \\
\hline
\hline
1 & 2,217 & 1,111 & 1,079 & 0.256\\
2 & 5,970 & 2,940 & 2,953 & 4.251\\
3 & 16,685 & 8,281 & 8,171 & 3,150 \\
4 & 35,096 & 17,320 & 17,264 & 380,220,376 \\
5 & 76,979 & 37,943 & 37,868 & $2.5e^{+22}$ \\
\hline
\end{tabular}
\caption{A simulation of a \Providence audit of a tied contest with an announced margin of 0.0566, and $\alpha=0.1$. Each round size is computed for a conditional stopping probability of $0.9$. The stopping condition is $\omega^{-1} \le \alpha$.}
\label{tab:toy}
\end{table}

@@

\npara{\bf The Adversary in an RLA:} The goal of the adversary is to increase the true risk beyond the declared risk limit---that is, given an incorrect outcome, make the audit declare it as correct with a chance larger than the risk limit. An audit is an RLA only if there is a proof of its risk-limiting property. Hence, at the very least, the adversary would need to invalidate an assumption of the proof to obtain:  
$$
Pr[\mathcal{A}(X) 
= \text{Correct} \,|\, H_0] > \alpha
$$

@@

The stopping condition for \Minerva does not take into account the sample obtained in previous rounds. The risk limit is estimated through weighted averages across multiple rounds, assuming that round sizes do not depend on the previous sample. 
%Further, it assumes that the null hypothesis is a tie. 
Attempting to simplify the proof of \Minerva's risk-limiting property, we were able to define a different audit \Providence. The RLA proof for \Providence does not make an assumption about round sizes.

@@

\subsubsection{Organization} 
Section \ref{sec:model} provides preliminaries on the \BRAVO and \Minerva audits. Section \ref{sec:prov} describes the \Providence audit, section \ref{sec:sims} the simulations comparing the number of ballots drawn using various ballot polling audits and section \ref{sec:pilot} the use of \Providence in an audit carried out by the Board of Elections of Rhode Island. Section \ref{sec:workload} presents our workload model and describes its use for a ballot polling audit using details of the 2016 US Presidential election in the state of Virginia, while section \ref{sec:misleading} introduces the notion of a misleading sample, illustrating it on the Virginia details as an example. Our conclusions, the availability of an audit implementation and acknowledgements may be found in sections \ref{sec:conc}, \ref{sec:avail} and \ref{sec:ack} respectively.

### 2023-02-28 providence/sections/related.tex
made the two changes from email
Changes to intro and bib

%Related Work
In this section, we describe related work.

\npara{\bf Ballot Polling RLA Process:} Bernhard provides a good description of the RLA and its assumptions, and also describes the process on the ground \cite{bernhard-sok}.

Election officials typically draw ballots in large round sizes, see for example \cite{va-2022,RI-report}. Note also that, in addition to allowing users to directly enter a round size or choose the expected number of ballots drawn by \BRAVO, Arlo provides choices of stopping probabilities of $0.9$, $0.8$ and $0.7$. For the two audits we attended, election officials chose stopping probabilities of at least $0.9$. Estimates of round sizes with stopping probability 0.9 for each state in the 2020 US Presidential election may be found in \cite{usenix_minerva}. Thousands of ballots is quite a common estimate; many estimates are as large as tens and hundreds of thousands of ballots. We are not aware of any ballot polling RLA performed on ballots cast in a governmental election that drew ballots one at a time (though the stopping condition can be computed one ballot at a time, the ballots are drawn in rounds).

\npara{\bf R2 and B2 Audits and the Classical \BRAVO audit:} A {\em round-by-round (R2)} audit is the general audit, where the decision of whether to draw more ballots or not is taken after drawing a round of ballots. A {\em ballot-by-ballot (B2)} audit is the special case of round size one---when the decision is made after each ballot is drawn. The popular \BRAVO audit requires the smallest expected number of ballots when the announced tally of the election is correct, and stopping decisions are taken a ballot at a time (that is, when it is used as a B2 audit). However, \BRAVO cannot be used as a B2 audit in the scenarios described in the previous paragraph.

@@

%The \BRAVO audit \cite{bravo} is the most popular ballot polling audit. When ballots are sampled one at a time, it is the audit with the smallest expected number of ballots drawn. 
\npara{\bf Newer Ballot Polling Audits:} The \Minerva audit \cite{usenix_minerva,arxiv_athena} does not need ballot order and relies only on sample and round tallies. It was developed for use with large first round sizes, and has been proven to be risk limiting when the round schedule for the audit is fixed before any ballots are drawn. First-round sizes for a stopping probability of $0.9$ when the announced tally is correct have been shown to be smaller than those for EoR and SO \BRAVO for a wide range of margins.

@@

Zag\'{o}rski {\em et al.}  present first-round simulations demonstrating that \Minerva draws fewer ballots than SO \BRAVO in the first round for large first round sizes when the true tally is as announced. Broadrick {\em et al.} provide further simulations showing that \Minerva requires fewer ballots than EoR and SO \BRAVO over multiple rounds and for smaller stopping probability. As expected, the advantage of \Minerva decreases for smaller stopping probability (smaller round sizes) as it approaches a B2 audit, 
%round schedule (1,1,1,\ldots) 
for which \BRAVO is known to be most efficient.

>>>>

%Related Work
\npara{\bf Ballot Polling RLA Process:} Bernhard provides a good description of the RLA and its assumptions, and also describes the process on the ground \cite{bernhard-sok}.

Election officials typically draw ballots in large round sizes, see for example \cite{va-2022,RI-report}. Note also that, in addition to allowing users to directly enter a round size or choose the expected number of ballots drawn by \BRAVO, Arlo provides choices of stopping probabilities of $0.9$, $0.8$ and $0.7$. For the two audits we attended, election officials chose stopping probabilities of $0.9$ and $0.95$. Estimates of round sizes with stopping probability 0.9 for each state in the 2020 US Presidential election may be found in \cite{usenix_minerva}. Thousands of ballots is quite a common estimate; many estimates are as large as tens and hundreds of thousands of ballots. We are not aware of any ballot polling RLA performed on ballots cast in a governmental election that drew ballots one at a time (though the stopping condition can be computed one ballot at a time, the ballots are drawn in rounds).

\npara{\bf R2 and B2 Audits and the Classical \BRAVO audit:} A {\em round-by-round (R2)} audit is the general audit, where the decision of whether to draw more ballots or not is taken after drawing a round of ballots. A {\em ballot-by-ballot (B2)} audit is the special case of round size one---when the decision is made after each ballot is drawn. The popular \BRAVO audit \cite{bravo} requires the smallest expected number of ballots when the announced tally of the election is correct, and stopping decisions are taken a ballot at a time (that is, when it is used as a B2 audit). However, \BRAVO cannot be used as a B2 audit in the scenarios described in the previous paragraph.

@@

%The \BRAVO audit \cite{bravo} is the most popular ballot polling audit. When ballots are sampled one at a time, it is the audit with the smallest expected number of ballots drawn. 
\npara{\bf Newer Ballot Polling Audits:} The \Minerva audit \cite{usenix_minerva,arxiv_athena} does not need ballot order and relies only on sample and round tallies. It was developed for use with large first round sizes, and has been proven to be risk limiting when the round schedule for the audit is fixed before any ballots are drawn. First round sizes for a stopping probability of $0.9$ when the announced tally is correct have been shown to be smaller than those for EoR and SO \BRAVO for a wide range of margins.

@@

Zag\'{o}rski {\em et al.}  present first round simulations demonstrating that \Minerva draws fewer ballots than SO \BRAVO in the first round for large first round sizes when the true tally is as announced. Broadrick {\em et al.} provide further simulations showing that \Minerva requires fewer ballots than EoR and SO \BRAVO over multiple rounds and for smaller stopping probability. As expected, the advantage of \Minerva decreases for smaller stopping probability (smaller round sizes) as it approaches a B2 audit, 
%round schedule (1,1,1,\ldots) 
for which \BRAVO is known to be most efficient.

### 2023-02-28 providence/sections/sims.tex
Changes to intro and bib

\begin{figure}[h!]
\includegraphics[width=.5\textwidth]{prov_risk.png}
\caption{The fraction of simulated \Providence audits on tied elections that stopped in any rounds (we performed five rounds at a risk limit of $0.1$) as a function of contest margin. This value is an estimate of the maximum risk of the \Providence audit.}
\label{fig:prov-risk}
\end{figure}

>>>>

\begin{figure}[h!]
\includegraphics[width=.5\textwidth]{prov_risk.png}
\caption{The fraction of simulated \Providence audits on tied elections that stopped in any round (we performed five rounds at a risk limit of $0.1$) as a function of contest margin. This value is an estimate of the maximum risk of the \Providence audit. Observe that it is below the risk limit, as expected.}
\label{fig:prov-risk}
\end{figure}
